---
title: "Examples"
output: arl::arl_html_vignette
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
arl::register_knitr_engine()
```

These examples are complete Arl programs from `inst/examples`. Each section
pulls in the full source from the corresponding file at build time, with
syntax highlighting, plus a link back to the repository version.

## Fibonacci

Recursive, iterative, and sequence-based Fibonacci implementations.

Source: <https://github.com/wwbrannon/arl/blob/main/inst/examples/fibonacci.arl>

```{r echo=FALSE, results='asis'}
path <- system.file("examples", "fibonacci.arl", package = "arl")
if (path == "") path <- file.path("..", "inst", "examples", "fibonacci.arl")
cat("```lisp\n")
cat(readLines(path, warn = FALSE), sep = "\n")
cat("\n```\n")
```


## Quicksort

Quicksort and mergesort, with list operations and helpers.

Source: <https://github.com/wwbrannon/arl/blob/main/inst/examples/quicksort.arl>

```{r echo=FALSE, results='asis'}
path <- system.file("examples", "quicksort.arl", package = "arl")
if (path == "") path <- file.path("..", "inst", "examples", "quicksort.arl")
cat("```lisp\n")
cat(readLines(path, warn = FALSE), sep = "\n")
cat("\n```\n")
```


## FizzBuzz

Multiple FizzBuzz variations showing control flow and list processing.

Source: <https://github.com/wwbrannon/arl/blob/main/inst/examples/fizzbuzz.arl>

```{r echo=FALSE, results='asis'}
path <- system.file("examples", "fizzbuzz.arl", package = "arl")
if (path == "") path <- file.path("..", "inst", "examples", "fizzbuzz.arl")
cat("```lisp\n")
cat(readLines(path, warn = FALSE), sep = "\n")
cat("\n```\n")
```


## Macro Examples

Macro definitions, quasiquote, expansion, and DSL-style helpers.

Source: <https://github.com/wwbrannon/arl/blob/main/inst/examples/macro-examples.arl>

```{r echo=FALSE, results='asis'}
path <- system.file("examples", "macro-examples.arl", package = "arl")
if (path == "") path <- file.path("..", "inst", "examples", "macro-examples.arl")
cat("```lisp\n")
cat(readLines(path, warn = FALSE), sep = "\n")
cat("\n```\n")
```


## Pipeline Macros

Threading a value through a custom pipeline macro.

Source: <https://github.com/wwbrannon/arl/blob/main/inst/examples/pipeline-macros.arl>

```{r echo=FALSE, results='asis'}
path <- system.file("examples", "pipeline-macros.arl", package = "arl")
if (path == "") path <- file.path("..", "inst", "examples", "pipeline-macros.arl")
cat("```lisp\n")
cat(readLines(path, warn = FALSE), sep = "\n")
cat("\n```\n")
```


## Data Analysis

R interoperability and data transformations.

Source: <https://github.com/wwbrannon/arl/blob/main/inst/examples/data-analysis.arl>

```{r echo=FALSE, results='asis'}
path <- system.file("examples", "data-analysis.arl", package = "arl")
if (path == "") path <- file.path("..", "inst", "examples", "data-analysis.arl")
cat("```lisp\n")
cat(readLines(path, warn = FALSE), sep = "\n")
cat("\n```\n")
```


## Graph Paths

BFS traversal and Dijkstra shortest paths.

Source: <https://github.com/wwbrannon/arl/blob/main/inst/examples/graph-paths.arl>

```{r echo=FALSE, results='asis'}
path <- system.file("examples", "graph-paths.arl", package = "arl")
if (path == "") path <- file.path("..", "inst", "examples", "graph-paths.arl")
cat("```lisp\n")
cat(readLines(path, warn = FALSE), sep = "\n")
cat("\n```\n")
```


## Log Parser

Parsing logs, aggregating status counts, and computing averages.

Source: <https://github.com/wwbrannon/arl/blob/main/inst/examples/log-parser.arl>

```{r echo=FALSE, results='asis'}
path <- system.file("examples", "log-parser.arl", package = "arl")
if (path == "") path <- file.path("..", "inst", "examples", "log-parser.arl")
cat("```lisp\n")
cat(readLines(path, warn = FALSE), sep = "\n")
cat("\n```\n")
```


## Sales Report

Data wrangling and CSV output using R interop.

Source: <https://github.com/wwbrannon/arl/blob/main/inst/examples/sales-report.arl>

```{r echo=FALSE, results='asis'}
path <- system.file("examples", "sales-report.arl", package = "arl")
if (path == "") path <- file.path("..", "inst", "examples", "sales-report.arl")
cat("```lisp\n")
cat(readLines(path, warn = FALSE), sep = "\n")
cat("\n```\n")
```


## Task Runner

Dependency ordering and execution sequencing.

Source: <https://github.com/wwbrannon/arl/blob/main/inst/examples/task-runner.arl>

```{r echo=FALSE, results='asis'}
path <- system.file("examples", "task-runner.arl", package = "arl")
if (path == "") path <- file.path("..", "inst", "examples", "task-runner.arl")
cat("```lisp\n")
cat(readLines(path, warn = FALSE), sep = "\n")
cat("\n```\n")
```

