---
title: "Getting Started"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Rye is a Lisp dialect implemented in and with access to R. This vignette
walks through installing Rye, running the REPL, and evaluating a few
expressions.

## Installation

```{r eval=FALSE}
devtools::install()
```

## Start the REPL

```{r eval=FALSE}
library(rye)
rye_repl()
```

At the prompt, you can enter Rye expressions:

```lisp
(+ 1 2)
(define x 10)
(* x 2)
```

## Run from the command line

If you installed the CLI wrapper, you can evaluate code without opening R:

```bash
rye --eval "(+ 1 2)"
rye --file script.rye
```

You can also pass multiple files; they run in order.

## Evaluate Rye from R

If you want to evaluate Rye inside an R script, you can parse and evaluate
expressions directly:

```{r eval=FALSE}
exprs <- rye_read("(define x 10) (+ x 5)")
rye_eval(exprs[[1]])
rye_eval(exprs[[2]])
```

If you have multiple expressions, `rye_eval_exprs()` will evaluate them
sequentially and return the last value.

## Run Rye files

You can execute a file from the REPL using `load`:

```lisp
(load "script.rye")
```

From R, use `rye_load_file()`:

```{r eval=FALSE}
rye_load_file("script.rye")
```

## Loading stdlib modules

The REPL loads the base stdlib. You can load extra modules from `inst/rye`:

```lisp
(load "control")
(load "binding")
(load "looping")
(load "threading")
(load "error")
```

## Core syntax

```lisp
; Function definition
(define factorial
  (lambda (n)
    (if (< n 2)
      1
      (* n (factorial (- n 1))))))

; Lambda destructuring (pattern wrapper keeps defaults unambiguous)
(define sum-pair
  (lambda ((pattern (a b)))
    (+ a b)))

; Dotted rest destructuring
(define head-tail
  (lambda (x . (pattern (a b)))
    (list x a b)))

; Lists
(list 1 2 3)
(car (list 1 2 3))
(cdr (list 1 2 3))

; Quoting
'(+ 1 2)           ; => (+ 1 2)
`(list ,x 20 30)   ; => (list 10 20 30)

; Comments start with semicolon
; This line is ignored
```

For more details on the standard library and macros, see the other vignettes in
this package.

## Next steps

- [Macros and Quasiquote](macros.html)
- [R Interop and Data Workflows](r-interop.html)
- [Standard Library Overview](stdlib-reference.html)
